---
title: "PM 566: Lab 04"
author: "Tarun Mahesh"
format: 
  html:
    embed-resources: true
fig-width: 10
fig-height: 10
theme: sandstone
editor: 
  markdown: 
    wrap: 72
---

STEPS

1.  Loading in the met dataset and reading it. I have already previously
    downloaded the dataset to my github repo (.gitignore) and will thus
    not be downloading it again.

```{r}
library(dplyr)
met <- read.csv("data/met_all.gz")
```

2.  Data Pre-Prep

    Things to do:

    -   Remove temperatures less than -17C
    -   Make sure there is no missing data in the key variables coded as
        9999, 999, etc.
    -   Generate a date variable using the functions as.Date() (hint:
        You will need the following to create a date paste(year, month,
        day, sep = "-")).
    -   Subset the data to keep only the observations from the first
        week (ie. first 7 days) of the month.
    -   Compute the mean by station of the variables temp, rh, wind.sp,
        vis.dist, dew.point, lat, lon, and elev.
    -   Create a region variable for NW, SW, NE, SE based on lon = and
        lat = degrees
    -   Create a categorical variable for elevation as in the lecture
        slides

```{r}
met <- met[met$temp > -17 & !is.na(met$temp), ]
met$elev[met$elev == 9999.0] <- NA
met$week <- as.numeric(format(as.Date(paste(met$year, met$month, 
                                            met$day, sep = "-")), "%U"))

met <- met[met$week == min(met$week, na.rm = TRUE), ]
met_avg <- aggregate(cbind(temp, rh, wind.sp, vis.dist, dew.point, 
                           lat, lon, elev, atm.press) ~ USAFID, 
                     data = met, FUN = mean, na.rm = TRUE)

met_avg$region <- ifelse(met_avg$lon > -98 & met_avg$lat > 39.71, "north east", 
                         ifelse(met_avg$lon > -98 & met_avg$lat < 39.71, 
                                "south east", ifelse(met_avg$lon < -98 & 
                                                       met_avg$lat > 39.71, 
                                                     "north west", 
                                                     "south west")))

met_avg$elev_cat <- ifelse(met_avg$elev > 252, "high", "low")

```

3.  Use geom_violin to examine the relative humidity and dew point by
    region
    -   Use facets
    -   Deal with NAs
    -   Describe observations

```{r}
library(ggplot2)

met_avg %>%
  filter(!(region %in% NA)) %>%
ggplot() + facet_wrap(~region, nrow = 2)

met_avg %>%
  filter(!(dew.point %in% NA)) %>%
ggplot() + geom_violin(mapping = aes(y = dew.point, x = 1)) + 
  facet_wrap(~region, nrow = 2)

met_avg %>%
  filter(!(dew.point %in% NA)) %>%
ggplot() + geom_violin(mapping = aes(y = wind.sp, x = 1)) + 
  facet_wrap(~region, nrow = 2)

met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() + geom_boxplot(mapping = aes(y = rh, fill = region)) + 
  facet_wrap(~region, nrow = 2)
```

Plot description:

The violin plots reveal distinct regional patterns in wind speed
distributions. The northeast shows a symmetric distribution, with most
values clustered around 3 m/s, while the northwest displays greater
variability. The southeast exhibits less variability, whereas the
southwest demonstrates wind speeds with the greatest variation compared
to the other regions. Overall, the northeast region appears to have the
highest wind speeds, while the southwest region shows the most
concentrated distribution around 3 m/s.

4.  Use geom_jitter with stat \_smooth to examine the association
    between dew point and wind speed by region
    -   Deal with NAs
    -   Color points by region
    -   Fit Linear Regression by Region

Plot Description:

The scatter plot illustrates the relationship between dew point and
relative humidity across all regions. All regional regression lines
exhibit positive slopes, indicating that relative humidity increases as
the dew point rises. The relationship appears strongest in the Northwest
and Northeast regions, while the Southeast and Southwest show a more
moderate association.

```{r}
# Dew Point and Relative Humidity
met_avg %>%
  filter(!(dew.point %in% NA)) %>% 
ggplot(mapping = aes(x = dew.point, y = rh, colour = region)) + 
geom_jitter() + stat_smooth(method = lm)
```

```{r}
#Dew Point and Wind Speed
met_avg %>%
  filter(!(dew.point %in% NA)) %>% 
  ggplot(mapping = aes(x = dew.point, y = wind.sp, colour = region)) + 
  geom_jitter() +
  stat_smooth(method = lm) 
```

```{r}
#Removing NA values using !is.na for both plots 
met_avg %>%
filter(!is.na(dew.point) & !is.na(rh)) %>% 
  ggplot(mapping = aes(x = dew.point, y = rh, colour = region)) + 
  geom_jitter() + stat_smooth(method = lm) 

met_avg %>%
filter(!is.na(dew.point) & !is.na(wind.sp)) %>% 
  ggplot(mapping = aes(x = dew.point, y = wind.sp, colour = region)) + 
  geom_jitter() + stat_smooth(method = lm) 
```

5.  Use geom_bar to create barplots of the weather stations by elevation
    category colored by region

    -   Deal with NAs

    -   Plot bars by category "elevation" using position = "dodge"

    -   Change colors from the default color by region using
        scale_fill_brewer

    -   Add labels and titles

    Plot Description:

    The bar chart highlights differences in weather station distribution
    across regions and elevation categories. The Southeast region
    contains highest number of stations overall, most of which are at
    low elevations. The Northeast region has greater number of stations
    at higher elevations. The Northwest region has fewest stations in
    total but shows a high proportion at elevated locations. The
    Southwest region falls in between, with moderate number of stations,
    many are also located at higher elevations.

```{r}
met_avg %>%
  filter(!(elev_cat %in% NA)) %>%
  ggplot() +
geom_bar(mapping = aes(x = elev_cat, fill = region), position = "dodge") +
  scale_fill_brewer(palette = "PuOr") + 
  labs(title = "Number of Weather Station sby Elevation Category and Region", x = "Elevation Category", y = "Count") + theme_bw()
```

6.  Use stat_summary to examine mean dew point and wind speed by region
    with standard deviation and error bars.

    Plot Description:

    Southeast region exhibits the highest mean dew point, while
    Northeast has the lowest. The error bars indicate that Southwest
    experiences the greatest variability, reflecting wider range of dew
    point values compared to other regions. When comparing wind speed
    patterns with dew point distributions, the Northwest and Southwest
    regions show higher mean wind speeds, around 3 m/s, suggesting
    greater variability in wind conditions. In contrast, the Northeast
    and Southeast regions display lower mean wind speeds with less
    variability.

```{r}
# Dew Point by Region
met_avg %>%
  filter(!is.na(dew.point) & !is.na(region)) %>%
  ggplot(mapping = aes(x = region, y = dew.point)) +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar") +
  stat_summary(fun.data = "mean_sdl")
```

```{r}
# Wind Speed by Region
met_avg %>%
  filter(!is.na(wind.sp) & !is.na(region)) %>%
  ggplot(mapping = aes(x = region, y = wind.sp)) +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar") +
  stat_summary(fun.data = "mean_sdl")
```

7.  Make a map showing the spatial trend in relative humidity in the US

    Plot Description:

    Relative humidity map reveals distinct east–west gradient across the
    United States. The eastern regions have higher relative humidity
    values while the central regions show moderate levels and the
    western regions exhibit lower values. The ten locations with the
    highest relative humidity are concentrated primarily in the eastern
    and southeastern United States.

```{r}
library(leaflet)
met_avg2 <- met_avg[!is.na(met_avg$rh), ]

top10 <- met_avg2[rank(-met_avg$rh) <= 10, ]

rh_pal = colorNumeric(c('blue', 'purple', 'red'), domain = met_avg2$rh)

leaflet(met_avg2) %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lat =~ lat, lng =~ lon, color =~ rh_pal(rh), 
             label =~ paste0(round(rh, 2), 'rh'), opacity = 1, fillOpacity = 1, radius = 500) %>%
  addMarkers(lat =~ lat, lng =~ lon, label =~ paste0(round(rh, 2), 'rh'), data = top10) %>%
  addLegend('bottomleft', pal = rh_pal, values = met_avg2$rh, 
            title = "Relative Humidity", opacity = 1)
  
```

8.  Use a ggplot extension to make a plot of our choosing

    I would like to analyse the atmospheric pressure variations by
    region, and I shall use the ridgeline extension

    ```{r}
    library(ggridges)

    met_avg %>% 
      filter(!(temp %in% NA)) %>%
    ggplot(mapping = aes(x = atm.press, y = region, fill = region)) +
    geom_density_ridges(alpha = 0.7) +
    facet_wrap(~elev_cat, nrow = 2) +
    scale_fill_brewer(palette = "Set2") +
    labs(title = "Atmospheric Pressure Distributions by Region",
    x = "Atmospheric Pressure", y = "Region", fill = "Region") +
    theme_bw() +
    theme(legend.position = "bottom")
    ```

    Description:

    These ridge plots show regional differences in atmospheric pressure
    under high and low conditions. The Northeast and Northwest cluster
    around higher pressures, while the Southeast and Southwest display
    more variable distributions. The Southwest,peaks lower under low
    pressure conditions, showing stronger variability compared to other
    regions.
