---
title: "PM 566: Lab 09"
author: "Tarun Mahesh"
execute:
  warning: false
format: 
  html:
    embed-resources: true
fig-width: 10
fig-height: 10
theme: sandstone
---

# SETUP

```{r}
# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
actor <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/actor.csv")
rental <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/rental.csv")
customer <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/customer.csv")
payment <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/payment_p2007_01.csv")

# Copy data.frames to database
dbWriteTable(con, "actor", actor)
dbWriteTable(con, "rental", rental)
dbWriteTable(con, "customer", customer)
dbWriteTable(con, "payment", payment)
```

```{r}
dbListTables(con)
```

# **Exercise 1**

Edit the code below to retrieve the actor ID, first name and last name for all actors using the `actor` table. Sort by last name and then by first name

```{sql connection=con}
SELECT actor_id, first_name, last_name
FROM actor
ORDER BY last_name, first_name;
```

# Exercise 2

Actors with last name WILLIAMS or DAVIS.

```{sql connection=con}
SELECT actor_id, first_name, last_name
FROM actor
WHERE last_name IN ('WILLIAMS', 'DAVIS');
```

# Exercise 3

Customer IDs who rented on 2005-07-05 (distinct).

```{sql connection=con}
SELECT DISTINCT customer_id
FROM rental
WHERE date(rental_date) = '2005-07-05';
```

# Exercise 4

## 4.1

Rows in `payment` where amount is 1.99, 7.99, or 9.99.

```{sql connection=con}
SELECT *
FROM payment
WHERE amount IN (1.99, 7.99, 9.99);
```

## 4.2

Rows where amount \> 5.

```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 5;
```

## 4.3

Rows where amount \> 5 and \< 8.

```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 5 AND amount < 8;
```

# Exercise 5

Payment IDs and amounts for customers with last name ‘DAVIS’.

```{sql connection=con}
SELECT payment_id, amount
FROM payment AS p
INNER JOIN customer AS c
  ON p.customer_id = c.customer_id
WHERE last_name IN ('DAVIS')
```

# Exercise 6

## 6.1

Use `COUNT(*)` to count the number of rows in `rental`.

```{sql connection=con}
SELECT COUNT(*) 
FROM rental
```

## 6.2

Use `COUNT(*)` and `GROUP BY` to count the number of rentals for each `customer_id`.

```{sql connection=con}
SELECT customer_id, COUNT(*) 
FROM rental
GROUP BY customer_id
```

## 6.3

Repeat the previous query and sort by the count in descending order.

```{sql connection=con}
SELECT customer_id, COUNT(*) AS n_rentals
FROM rental
GROUP BY customer_id
ORDER BY n_rentals DESC
```

## 6.4

Repeat the previous query but use `HAVING` to only keep the groups with 40 or more.

```{sql connection=con}
SELECT customer_id, COUNT(*) AS n_rentals
FROM rental
GROUP BY customer_id
HAVING COUNT(*) >= 40
ORDER BY n_rentals DESC
```

# Exercise 7

Write a query that calculates a number of summary statistics for the payment table using `MAX`, `MIN`, `AVG` and `SUM`

```{sql connection=con}
SELECT
  MAX(amount),
  MIN(amount),
  AVG(amount),
  SUM(amount)
FROM payment
```

## 7.1

Modify the above query to do those calculations for each `customer_id`.

```{sql connection=con}
SELECT
  customer_id,
  MAX(amount),
  MIN(amount),
  AVG(amount),
  SUM(amount),
  COUNT(*)   
FROM payment
GROUP BY customer_id
```

## 7.2

Modify the above query to only keep the `customer_id`s that have more then 5 payments.

```{sql connection=con}
SELECT
  customer_id,
  MAX(amount),
  MIN(amount),
  AVG(amount),
  SUM(amount),
  COUNT(*)    
FROM payment
GROUP BY customer_id
HAVING COUNT(*) > 5
```

# CLEANUP

```{r}
dbDisconnect(con)
```
